---
title: 'BFI Orientation 2022: Introduction to R'
output:
  html_notebook: default
  html_document:
    df_print: paged
  pdf_document: default
---


## Data visualization with ggplot2 
The `ggplot2` package is based on the **grammar of graphics**, the idea that you can build every graph from the same few components: a data set, a set of geoms (data points), and a coordinate system.

load the package
```{r library, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2) 
library(scales) # breaks and labels for axes and legends
library(haven) # read_dta
library(stargazer) # export regression table
library(starpolishr) # complementary package of stargazer
library(lfe) # fixed effect regression
library(fixest) # (fast) fixed effect regression
library(modelsummary) # get estimated model from regression
library(stringr) # some useful string functions
library(stats) # some statistical calculations and random number generation.
library(ggsignif) # add mean comparison with significant values in graph
```

some ggplot basic examples
```{r ggplot-basic, message=FALSE, warning=FALSE}
# import data
df <- read.csv("data/tv_hurricanes_by_network.csv")
plot_df <- gather(df, 
        key = "tv.station", #name of the new column of 'old columns to be gathered'
        value = "coverage", #name of the 'gathered values' 
        BBC.News:MSNBC) 

#create density plots  
ggplot(data = plot_df) +
  geom_density(aes(x = coverage, fill = tv.station), alpha = 0.3) + # can also change the fill to color
  theme_classic()  +
  xlab("news coverage") #label the x axis

#create a timeseries of news coverage 
ggplot(data = plot_df) +
  geom_line(aes(x = as.Date(Date, "%m/%d/%y"), y = coverage, 
                group = tv.station, 
                color = tv.station), alpha = 0.7) +
  xlab("date") + 
  scale_x_date(labels = date_format("%m/%d"),
               breaks = date_breaks("1 week"))   +
  theme_classic()

#we could also have separate plots by tv station by using facet_wrap()
ggplot(data = plot_df) +
  geom_line(aes(x = as.Date(Date, "%m/%d/%y"), y = coverage, 
                group = tv.station, 
                color = tv.station), alpha = 0.5) +
  facet_wrap(~tv.station, ncol = 4) + 
  xlab("date") + 
  scale_x_date(labels = date_format("%m/%d"),
               breaks = date_breaks("2 weeks"))   +
  theme_classic()

#create histograms of each hurricane in terms of number of observations
ggplot(data = plot_df) +
  geom_bar(aes(x = Query, color = Query), 
           color = "black",
           stat = "count", #calculates the number of cases in each group
           alpha = 0.3)  +
  theme_classic()

#create scatterplot of news coverage against time by TV Station
ggplot(data = plot_df) +
  geom_point(aes(y = coverage, x = as.Date(Date, "%m/%d/%y"), color = tv.station),
             shape = 16,
             size = 4,
             alpha = 0.3) + 
  xlab("date") +  
  theme_classic() +
  scale_x_date(labels = date_format("%m/%d"),
               breaks = date_breaks("1 week")) 


#overlay scatterplot on timeseries lineplot
ggplot(data = plot_df) +
  geom_line(aes(x = as.Date(Date, "%m/%d/%y"), y = coverage, 
                group = tv.station, 
                color = tv.station), alpha = 0.7) +
  geom_point(aes(y = coverage, x = as.Date(Date, "%m/%d/%y"), color = tv.station),
             shape = 16,
             size = 4,
             alpha = 0.3) + 
  xlab("date") + 
  theme_classic() +
  scale_x_date(labels = date_format("%m/%d"),
               breaks = date_breaks("1 week")) 

```


## Regression and Table Output 
- linear regression `lm` and the formula
- table output: using `stargazer`
- introduction to fast fixed-effects `fixest`
- more table output: summary stats

- Example from: Sanchez de la Sierra, R. (2021). Whither Formal Contracts?. Econometrica, 89(5), 2341-2373. https://www.econometricsociety.org/publications/econometrica/2021/09/01/whither-formal-contracts.

linear regression and the formula
```{r lm, message=FALSE, warning=FALSE}
data = read_dta('data/ANALYSIS.dta') %>%
  filter(B1 == 1) 

controls_contract = 'REPLACE_HH+HH_OCCUPATION_PUBLIC+HOUSEHOLD_STATE'

model1 = lm(as.formula(str_interp('E~T_contract')), data=data)
model2 = lm(as.formula(str_interp('E~T_contract')), data=data %>% filter(STATE == 1))
model3 = lm(as.formula(str_interp('E~T_contract')), data=data %>% filter(STATE == 0))

model4 = lm(as.formula(str_interp('E~T_contract+equipe+q4_avenue_name+${controls_contract}')), 
                       data=data)
model5 = lm(as.formula(str_interp('E~T_contract+equipe+q4_avenue_name+${controls_contract}')), 
                       data=data %>% filter(STATE == 1))
model6 = lm(as.formula(str_interp('E~T_contract+equipe+q4_avenue_name+${controls_contract}')), 
                       data=data %>% filter(STATE == 0))
```

one basic example of stargazer
```{r stargazer, message=FALSE, warning=FALSE}
mean = formatC(mean(data$E[data$T_contract==0], na.rm = T), digits=3, format='f')
mean_list = str_interp(paste(rep('${mean}', 6), collapse = ' & '))
mean_list = str_interp('No Contract mean & ${mean_list} \\\\')

out = stargazer(model1, model2, model3, model4, model5, model6,
                keep = 'T_contract',
                covariate.labels = c('Contract'),
                keep.stat = c('rsq','n'), 
                dep.var.caption = 'Pay, conditional on initially accept',
                dep.var.labels.include = FALSE,
                title = 'Table 1', 
                label = 't:contract',
                add.lines = list(c('Sample', 'Full', 'Favored', 'Disfavored','Full', 'Favored', 'Disfavored'),
                                 c('Controls', rep('No',3), rep('Yes',3))))

# add no control mean
obs_row = which(str_detect(out, 'Observations'))

out = c(out[1:obs_row], mean_list, out[(obs_row+1):length(out)])
out = star_notes_tex(out, note.type = 'threeparttable', note = '') 

writeLines(out, con = 'output/example-1.tex')
```

introduction to fixest
```{r fixest, message=FALSE, warning=FALSE}
model1 = fixest::feols(as.formula(str_interp('E~T_contract')), data=data)
model2 = fixest::feols(as.formula(str_interp('E~T_contract')), data=data %>% filter(STATE == 1))
model3 = fixest::feols(as.formula(str_interp('E~T_contract')), data=data %>% filter(STATE == 0))

model4 = fixest::feols(as.formula(str_interp('E~T_contract|equipe+q4_avenue_name+${controls_contract}')), 
                       data=data)
model5 = fixest::feols(as.formula(str_interp('E~T_contract|equipe+q4_avenue_name+${controls_contract}')), 
                       data=data %>% filter(STATE == 1))
model6 = fixest::feols(as.formula(str_interp('E~T_contract|equipe+q4_avenue_name+${controls_contract}')), 
                       data=data %>% filter(STATE == 0))

models = list(model1, model2, model3, model4, model5, model6)

tex = modelsummary(models,
                   stars = c('*' = .1, '**' = .05, '***' = .01),
                   gof_omit = 'Adj|Within|AIC|BIC|Log',
                   output = 'latex',
                   escape = F) 
```

create a summary-stats table
```{r summary-stats, message=FALSE, warning=FALSE}
make_table = function() {
  
  data = plot_df %>%
    spread(key = Query, value = coverage)
  
  tag  ='sum-stat'
  path = str_interp('output/example-2.tex')
  digits = 2
  
  make_row = function(outcome) {
    
    dict = c('Hurricane Harvey'='Harvey',
             'Hurricane Irma'='Irma',
             'Hurricane Jose'='Jose',
             'Hurricane Maria'='Maria')
    
    data$outcome = data[[outcome]]

    values1 = c(mean(data$outcome[data$tv.station == 'BBC.News'], na.rm=T),
                mean(data$outcome[data$tv.station == 'CNN'],na.rm=T), 
                mean(data$outcome[data$tv.station == 'FOX.News'], na.rm=T),
                mean(data$outcome[data$tv.station == 'MSNBC'], na.rm=T))
    
    values1 = c(dict[outcome], map_chr(values1, function(x) formatC(x, digits=digits, format='f')))
    
    values2 = c(sd(data$outcome[data$tv.station == 'BBC.News'], na.rm=T),
                sd(data$outcome[data$tv.station == 'CNN'],na.rm=T), 
                sd(data$outcome[data$tv.station == 'FOX.News'], na.rm=T),
                sd(data$outcome[data$tv.station == 'MSNBC'], na.rm=T))
    
    values2 = c(dict[outcome], map_chr(values2, function(x) formatC(x, digits=digits, format='f')))
    values2=values2[2:length(values2)]

    values1 = paste(values1, collapse = ' & ')
    values1 = paste(values1, '\\\\')
    
    values2 = paste(values2, collapse = ') & (')
    values2 = paste0(' &(', values2, ')', '\\\\')

    values = paste(values1, values2)
    
    return(values)
  }
  
  rows = purrr::map(c( 'Hurricane Harvey',
                       'Hurricane Irma',
                       'Hurricane Jose',
                       'Hurricane Maria'), make_row)

  obs = c('Observations', 
          nrow(data %>% filter(tv.station == 'BBC.News')), 
          nrow(data %>% filter(tv.station == 'CNN')), 
          nrow(data %>% filter(tv.station == 'FOX.News')),
          nrow(data %>% filter(tv.station == 'MSNBC')))
  
  obs = paste(obs, collapse = ' & ')
  obs = paste(obs, '\\\\')
  
  
  out_put = c('\\begin{table}[H]',
              '\\centering', 
              '\\caption{Summary statistics}',
              str_interp('\\label{t:${tag}}'),
              '\\begin{threeparttable}',
              '\\begin{tabular}{lcccc}',
              '\\toprule',
              '& BBC & CNN & FOX & MSNBC \\\\',
              '\\midrule',
              unlist(rows), 
              '\\midrule',
              obs, 
              '\\bottomrule',
              '\\end{tabular}',
              '\\begin{tablenotes} \\footnotesize',
              '\\item \\textit{Notes:} ', # add notes
              '\\end{tablenotes}',
              '\\end{threeparttable}',
              '\\end{table}')
  print(out_put)
  
  writeLines(out_put, str_interp('${path}'))
  
}

make_table()
```


## Combine above 
- Create a bar with treatment effects, p-values, and some labels inside.

write some customized thing used in ggplot (palette, theme)
```{r function, message=FALSE, warning=FALSE}
p_format = function(x) {
  value = formatC(x, digits=3)
  if (x<0.001) {
    return(str_interp('p<0.001'))
  } else if (x>=0.001) {
    return(str_interp('p=${value}'))
  } 
}

#palette = c('#808080', '#bb0a1e', '#006400','#bb0a1e', '#006400')
palette = c('#808080', 'gray50', 'gray10','gray50', 'gray10')

theme_contract = list(theme_bw(base_size=12, base_family="LM Roman 10"),
                     theme(axis.title.x = element_blank(),
                           axis.title.y = element_text(size=12),
                           axis.text = element_text(size=12)),
                     guides(fill=F,col=F))

four_palette = list(scale_fill_manual(values=palette[2:5]),
                   scale_color_manual(values=palette[2:5]))
```

to get the p-values 
```{r get-pvalue, message=FALSE, warning=FALSE}
data = read_dta('data/ANALYSIS.dta') %>%
  filter(B1 == 1)

controls_contract = 'REPLACE_HH+HH_OCCUPATION_PUBLIC+HOUSEHOLD_STATE'

model1 = fixest::feols(as.formula(str_interp('E~T_contract|equipe+q4_avenue_name+${controls_contract}')), 
                       data=data %>% filter(STATE == 1))

pvalues = summary(model1)$coeftable[,'Pr(>|t|)']
p1 = p_format(pvalues)
p1 = str_interp('State-favored merchants \n ${p1}')


model2 = fixest::feols(as.formula(str_interp('E~T_contract|equipe+q4_avenue_name+${controls_contract}')), 
                       data=data %>% filter(STATE == 0))

pvalues = summary(model2)$coeftable[,'Pr(>|t|)']
p2 = p_format(pvalues)
p2 = str_interp('Disfavored merchants \n ${p2}')
```

plot the treatment effect bar
```{r combine, message=FALSE, warning=FALSE}
summary = data %>% 
  filter(!is.na(T_contract)) %>%
  group_by(T_contract, STATE) %>% 
  summarise(mean = mean(E), se = sd(E)/sqrt(n()), value = formatC(round(mean, 3), digits=3, format='f')) %>%
  arrange(desc(STATE), T_contract) %>%
  ungroup() %>%
  mutate(group = as.character(row_number())) %>%
  mutate(T_contract = as.character(T_contract)) # for scale_color_manual

mean_pos = 0.02
ypos_bar = max(summary$mean)+max(summary$mean)/4
ymax = max(summary$mean)+max(summary$mean)/2

tl = 0.03
textsize=5

ggplot(summary, aes(x = group, y = mean, fill = group)) +
  geom_bar(stat='identity', position="dodge", alpha=0.65, width=0.5) + 
  geom_errorbar(aes(ymin = mean-1.96*se, ymax = mean+1.96*se,col = T_contract), position = 'dodge', alpha=1, width=0.5) +
  geom_signif(comparisons = list(c('1','2')), annotations=str_interp(p1), textsize = textsize, family = 'LM Roman 10', y_position = min(c(0.95, max(summary$mean)+max(summary$mean)/4)), tip_length = tl, vjust = -.1) +
  geom_signif(comparisons = list(c('3','4')), annotations=str_interp(p2), textsize = textsize, family = 'LM Roman 10', y_position = min(c(0.95, max(summary$mean)+max(summary$mean)/4)), tip_length = tl, vjust = -.1) +
  ylab('Pays, conditional on initially accept') +
  geom_label(aes(x=group, y = mean_pos, label=value), color = 'black', fill = "white", alpha = .3, family = 'LM Roman 10', size = 5, label.size = NA) +
  coord_cartesian(ylim=c(0, ymax)) +
  theme_contract +
  four_palette +
  geom_vline(xintercept=2.5, linetype='dashed', alpha = 0.5) +
  scale_x_discrete(labels=c('1'='No Contract', '2'='Contract', '3'='No Contract', '4'='Contract')) 
```


### Others
- IV: `ivreg` ivreg(y ~ x1 | x2 | z1 + z2) (the exogenous, endogenous, and instrumental variables, respectively.)
- Fixest: feols(y ~ x1 | x_endo_1 + x_endo_2 ~ x_exo_1 + x_exo_2)





